A comparative analysis of penalized linear mixed models in structured genetic data

### 2020-08-18 main.tex
Title; making DAG accents pink because girl power
Section 2 edits

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices \citep{thornton2015statistical}.  In other words, random genetic variation is likely to be correlated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \citep{barton2019population}. Although bias at individual loci may be small, this bias becomes magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \citep{barton2019population, peterson2019genome}.

@@

Population structure is broadly categorized based on whether it describes recent or ancient relatedness. Recent relatedness describes familial or \textbf{cryptic relatedness}. Ancient relatedness, or common ancestry, is generally characterized by large clusters of low levels of similarity. \anna{I need some sort of subjective cutoff for recent vs. ancient relatedness rather than describing them in terms of clusters... more details} \pb{I would suggest something along the lines of: cryptic relatedness means that only a few generations, maybe 3 or so, separate common ancestors} The presence of multiple distinct ancestry groups in a sample is known as \textbf{population stratification}.

Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. However, in this review we focus on methods to account for unobserved relatedness. Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations. This phenomenon is commonly described as \textbf{confounding due to population stratification or ancestry}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure itself does not confound the genotype-phenotype relationship in the classical sense. While population structure may directly affect observed allele frequencies, there must also exist some mechanism by which population stratification affects the phenotype, namely, the environment \citep{barton2019population, vilhjalmsson2012nature}.

% \anna{DAG here with only population stratification, observed snps, and phenotype ... maybe an arrow from population stratification to phenotype with a question mark (?)}

@@

New statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. Many of these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in Figure \ref{fig:ps_env_block}. These methods do not directly address the environmental component of confounding due to population structure. Nevertheless, they may be effective if they are able to indirectly block the causal pathway between the environment and observed SNPs.

\begin{figure}[H]
\centering
\begin{tikzpicture}
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observbed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -] (1) edge (2);
    \path[red, line width = 0.5] (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[color=red,pos=.5,mark=x](1,3)
\end{tikzpicture}
\caption{The causal pathway between the environment and observed SNPs may be indirectly blocked by many methods that attempt to reduce confounding due to population structure.}
\label{fig:ps_env_block}
\end{figure}

Whether or not such methods are effective at reducing confounding effects depends upon whether population structure can be accurately inferred, and if this is a good proxy for environmental effects. In the presence of unobserved heterogeneous environmental effects which are not well-approximated by inferred population structure, these effects will not confound SNP-phenotype associations. This is due to the absence of a causal pathway between population stratification and environment, as depicted in Figure \ref{fig:ps_env_block2}. In such a scenario, failing to account for unobserved environmental effects will result in greater phenotypic variability \anna{and as a result, greater SE in beta-hats... if we're talking about gls}, however, it will not bias SNP estimates \anna{cite something from causal inference here}. We will return to this idea and its mathematical justification in section \ref{sec:methods}.

\begin{figure}[H]
\centering
\begin{tikzpicture}
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observbed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -, red] (1) edge (2);
    \path (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[color=red,pos=.5,mark=x](1,2)
\end{tikzpicture}
\caption{The causal pathway between the environment and observed SNPs may be blocked if genetic information does not serve as an accurate proxy for it.}
\label{fig:ps_env_block2}
\end{figure}

\anna{need transition to bring it back to the first DAG for now...} Stratified analysis attempts to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their more fine-grained, subtle relatedness.

>>>>

One reason that population stratification warrants such concern in genetic studies is that due to ethnic and geographic segregation, population stratification tends to be associated with differing environmental exposures and cultural practices \citep{thornton2015statistical, browning2011population}.  In other words, random genetic variation is likely to be correlated with non-genetic factors that also influence the trait of interest, thereby confounding the genotype-phenotype relationship and leading to spurious associations and biased estimates of SNP effects. Biased estimates of this nature severely hinder prediction and understanding differences among populations \citep{barton2019population}. Although bias at individual loci may be small, this bias becomes magnified when aggregated across thousands of SNPs, as is done when calculating polygenic risk scores \citep{barton2019population, peterson2019genome}.

@@

Population structure is defined by the existence of allele frequency differences that define sub-populations and is driven by the combined effects of evolutionary processes such as genetic drift, migration, demographic history, and natural selection \citep{gibson2015primer, tibayrenc2017genetics}. Population structure is broadly categorized based on whether it describes recent or ancient relatedness. Ancient relatedness describes the presence of a common ancestor many generations previously. The presence of distinct ancestry groups with different allele frequencies in a sample is known as \textbf{population stratification}. Recent relatedness describes the sharing of a common ancestor only several generations previously. Pedigree-based methods may be used to explicitly model recent relatedness if familial relationships are known. In this absence of known familial relationships, recent relatedness is referred to as \textbf{cryptic relatedness.} In this review we focus on methods to account for unobserved relatedness such as population stratification and cryptic relatedness.

Throughout this paper, the terms \textbf{population structure} and \textbf{structured population} will be used to describe a sample of individuals in which population stratification, cryptic relatedness, or both are present. Population stratification in particular has been of great concern in genetic studies due to its potential to lead to spurious associations when population structure is associated with differences in both allele frequency and the trait or disease \citep{gibson2015primer}. This phenomenon is commonly described as \textbf{confounding due to population stratification}. The mechanism of this phenomenon warrants some discussion as it is often overlooked that population structure in and of itself does not confound the genotype-phenotype relationship. While population structure may directly affect observed allele frequencies, there must also exist some non-genetic mechanism by which population stratification affects the phenotype, namely, the environment \citep{barton2019population, vilhjalmsson2012nature}.

@@

New statistical methods to control for the effects of confounding due to population stratification have proliferated as computational advances have allowed for the analysis of genetic data from large cohorts. Many of these methods attempt to block the causal pathway between population structure and observed SNP data. This is illustrated by the blocked arrow in Figure \ref{fig:ps_env_block}. As the figure shows, these methods do not directly address the environmental component of confounding due to population structure. Nevertheless, they may be effective in controlling confounding by indirectly blocking the causal pathway between the environment and observed SNPs.

\begin{figure}[H]
\centering
\begin{tikzpicture}
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observbed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -] (1) edge (2);
    \path[magenta, line width = 0.5] (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[color=magenta,pos=.5,mark=x](1,3)
\end{tikzpicture}
\caption{The causal pathway between the environment and observed SNPs may be indirectly blocked by many methods that attempt to reduce confounding due to population structure.}
\label{fig:ps_env_block}
\end{figure}

Whether or not such methods are effective at reducing confounding effects depends upon whether population structure can be accurately inferred, and if population structure is a good proxy for environmental effects. In the presence of environmental heterogeneity which affects the phenotype of interest, and which is not well-approximated by inferred population structure, SNP-phenotype associations will not be confounded. This is due to the absence of a causal pathway between population stratification and environment, as depicted in Figure \ref{fig:ps_env_block2}. In such a scenario, failing to account for unobserved environmental effects is likely to result in greater residual variability, however, SNP effect estimates will be unbiased \citep{greenland1999causal}. We will return to this idea and its mathematical justification in section \ref{sec:methods}.

\begin{figure}[H]
\centering
\begin{tikzpicture}
    \node (1) at (0,0) {Population Stratification};
    \node (2) [right = of 1] {Environment};
    \node (3) [below = of 1] {Observbed SNPs};
    \node (4) [below = of 2] {Phenotype};
    \path[bidirected, -, magenta] (1) edge (2);
    \path (1) edge (3);
    \path (2) edge (4);
    \path (3) edge (4);
    \tkzMarkSegment[color=magenta,pos=.5,mark=x](1,2)
\end{tikzpicture}
\caption{The causal pathway between the environment and observed SNPs may be blocked if genetic information does not serve as an accurate proxy for it.}
\label{fig:ps_env_block2}
\end{figure}

Assuming for now that population structure is a good proxy for environmental effects, we return to our consideration of methods that attempt to control for confounding via the mechanism depicted in Figure \ref{fig:ps_env_block}. Such methods include LMMs, principal-component-based approaches, and stratified analyses. Stratified analyses refer to a general class of methods which attempt to infer subject membership within a discrete number of nonoverlapping subpopulations. Subsequent analyses are conducted within each subpopulation, and the subpopulation-specific results may then be combined via meta-analysis \citep{pritchard1999use, pritchard2000association}. By partitioning the overall sample size in this manner, the structured association method is particularly vulnerable to a loss of power. Additionally, it relies on the assumption that these subpopulations are distinct and nonoveralapping. This assumption may not accurately reflect the characteristics of a particular sample, such as one consisting of admixed populations, or recent relatedness. Considered in the context of our air quality example, rather than the existence of two subpopulations with exposure to either good or bad air quality, there may also be individuals who fall between these two extremes and are characterized by continuously varying relatedness and a gradient of air quality. By clustering those individuals within in one subpopulation or another we lose information about their more fine-grained, subtle relatedness.
